---
title: "ml prj"
author: "me"
date: "17/08/2020"
output: html_document
---

##Synopsis

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways.

The goal of this project is to predict the manner in which they did the exercise. This is the `classe` variable in the training set.

## Data description

The outcome variable is `classe`, a factor variable with 5 levels. For this data set, participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in 5 different fashions:

- exactly according to the specification (Class A)
- throwing the elbows to the front (Class B)
- lifting the dumbbell only halfway (Class C)
- lowering the dumbbell only halfway (Class D)
- throwing the hips to the front (Class E)

## Initial configuration

The initial configuration consists of loading some required packages and initializing some variables.

```{r configuration, echo=TRUE, results='hide'}
#loading packages
library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(randomForest)
library(gbm)
# Set seed for reproducability
set.seed(9999)
```

## Data loading
In this section the data is downloaded and processed. Some basic transformations and cleanup will be performed, so that `NA` values are omitted. Irrelevant columns such as `user_name`, `raw_timestamp_part_1`, `raw_timestamp_part_2`, `cvtd_timestamp`, `new_window`, and  `num_window` (columns 1 to 7) will be removed in the subset.

The `pml-training.csv` data is used to devise training and testing sets.
The `pml-test.csv` data is used to predict and answer the 20 questions based on the trained model.

```{r dataloading, echo=TRUE}

train_url <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
test_url  <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"


init_org_training_data <- read.csv(url(train_url))
init_org_testing_data <- read.csv(url(test_url))
```



## Data preprocessing
```{r preprocessing, echo=TRUE}

#Removing Variables which are having nearly zero variance.
non_zero_var <- nearZeroVar(init_org_training_data)


org_training_data <- init_org_training_data[,-non_zero_var]
org_testing_data <- init_org_testing_data[,-non_zero_var]

dim(org_training_data)

#Removing Variables which are having NA values. Our threshhold is 95%.
na_val_col <- sapply(org_training_data, function(x) mean(is.na(x))) > 0.95

org_training_data <- org_training_data[,na_val_col == FALSE]
org_testing_data <- org_testing_data[,na_val_col == FALSE]

#Removing variables which are non-numeric and hence will not contribute into our model. The very first 7 variables are of that kind only. Hence those needs to be removed from the datasets.
org_training_data <- org_training_data[,8:59]
org_testing_data <- org_testing_data[,8:59]

dim(org_training_data)
```


## Data partitioning
```{r partitioning, echo=TRUE}
#As per recommendation of the course __ Practical Machine Learning__ , we will be seggregating our org_training_data into 2 different parts, one is the training set (consisiting 60% of the total data) and test set (consisting 40% of the total data)

inTrain <- createDataPartition(org_training_data$classe, p=0.6, list=FALSE)
training <- org_training_data[inTrain,]
testing <- org_training_data[-inTrain,]

dim(training)
```

## Decision Tree Model
```{r decisiontree, echo=TRUE}
DT_modfit <- train(classe ~ ., data = training, method="rpart")
#Prediction in terms of Decision Tree Model
DT_prediction <- predict(DT_modfit, testing)
confusionMatrix(DT_prediction, testing$classe)
rpart.plot(DT_modfit$finalModel, roundint=FALSE)
#We can see that the prediction accuracy is 50% which is not upto the desired level.
```

## Random Forest Model
```{r randomforest, echo=TRUE}
RF_modfit <- train(classe ~ ., data = training, method = "rf", ntree = 100)
# prediction in terms of random forest
RF_prediction <- predict(RF_modfit, testing)
RF_pred_conf <- confusionMatrix(RF_prediction, testing$classe)
RF_pred_conf
#here is the plot
plot(RF_pred_conf$table, col = RF_pred_conf$byClass, 
     main = paste("Random Forest - Accuracy Level =",
                  round(RF_pred_conf$overall['Accuracy'], 4)))
#From the Confusion Matrix, we can clearly see that the prediction accuracy of Random Forest model is 99% which is satisfactory.
```

## Gradient Boosting model
```{r gbm, echo=TRUE}
GBM_modfit <- train(classe ~ ., data = training, method = "gbm", verbose = FALSE)
GBM_modfit$finalModel
#Prediction in terms of GBM Model
#GBM_prediction <- predict(GBM_modfit, testing, type = "class", n.trees = 5, type = link)
GBM_prediction <- predict(GBM_modfit, testing)

GBM_pred_conf <- confusionMatrix(GBM_prediction, testing$classe)
GBM_pred_conf
#here is the plot
plot(GBM_pred_conf$table, col = GBM_pred_conf$byClass, 
     main = paste("Gradient Boosting - Accuracy Level =",
                  round(GBM_pred_conf$overall['Accuracy'], 4)))
#From Gradient Boost Model, the prediction accuracy is 95% which is satisfactory.
```

## Now we need to see how each model has predicted the validation dataset across the classifications. ** We are not considering Decision Tree model as it didn’t reach the satisfactory prediction accuracy level. SO only Random Forest and Gradient Boosting methods are being compared.

```{r prediction, echo=TRUE}
#random forest
RF_pred_conf$overall
#GBM
GBM_pred_conf$overall
```

## Conclusion
After checking the Overall Statistics data, the Random Forest model has definitely more accuracy than GBM. Hence we will be selecting Random Forest model for final prediction from org_testing_data.

## Final Prediction- Applying selected model on the Test Data
```{r finalprediction, echo=TRUE}
Final_RF_prediction <- predict(RF_modfit, org_testing_data )
Final_RF_prediction
```

